Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2020-04-10 ms/spaceness_final.tex
adding final ms files

\title{Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data}

@@

\runningauthor{Battey \textit{et al.}}

@@

The assumption that populations are ``well-mixed'' has important implications for downstream inference of selection and demography. Methods based on the coalescent \citep{Kingman1982,wakeley2005coalescent} assume that the sampled individuals are a random draw from a well-mixed population
that is much larger than the sample \citep{wakeley2003gene}. The key assumption is that the individuals of each generation are \emph{exchangeable}, so that there is no correlation between the fate or fecundity of a parent and that of their offspring \citep{Huillet2011}. If dispersal or mate selection is limited by geographic proximity, this assumption can be violated in many ways. For instance, if mean viability or fecundity is spatially autocorrelated, then limited geographic dispersal will lead to parent--offspring correlations. Furthermore, nearby individuals will be more closely related than an average random pair, so drawing multiple samples from the same area of the landscape will represent a biased sample of the genetic variation present in the whole population \citep{Stadler2009}.% It has long been appreciated that this model misspecification subjects downstream inferences to bias, but the extent and nature of these effects remain largely uninvestigated.

@@

% The best-studied approaches to population genetics in continuous space were developed by \citet{Wright1942,Wright1943} and \citet{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. 
The degree to which genetic relationships are geographically correlated
depends on the chance that two geographically nearby individuals are close relatives
-- in modern terms, by the tension between
migration (the chance that one is descended from a distant location)
and coalescence (the chance that they share a parent).
A key early observation by \citet{Wright1946} is that this balance is often nicely summarized by the ``neighborhood size'',
defined in two dimensions to be $N_W = 4 \pi \rho \sigma^2$,
where $\sigma^2$ is one half of the mean squared parent--offspring distance
and $\rho$ is population density
(see \citet{Rousset1997} for further discussion of parameter definitions in one- and two-dimensional habitats) \llabel{axialdist}. 
This can be thought of as proportional to the average number of potential mates for an individual (those within distance $2 \sigma)$,
or the number of potential parents of a randomly chosen individual. Empirical estimates of neighborhood size vary hugely across species
-- even in human populations,
estimates range from 40 to over 5,000 depending on the population and method of estimation (Table~\ref{table:NStable}).
%\citet{Maruyama1972} found that the rate of decline in genetic diversity in a two-dimensional continuous population approaches the random mating expectation when $\rho\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.

@@

To avoid hard-to-evaluate approximations, we here used forward-time, individual-based simulations
across continuous geographical space.
The question of what regulates real populations has a long history and many answers
\citep[e.g.,][]{lloyd1967crowding,antonovics1980ecological,crawley1990population}, \llabel{ecolcites}
but it is clear that populations must at some point have density-dependent feedback on population size,
or else they would face eventual extinction or explosion.
In the absence of unrealistic global population regulation,
this regulation must be local,
and there are many ways to achieve this \citep{bolker2003spatial}.
In our simulations,
each individual's probability of survival is a decreasing function of local population density,
which shifts reproductive output towards low-density regions,
and produces total census sizes that fluctuate around an equilibrium. 
This also prevents the population clumping seen by \citet{Felsenstein1975}
(Supplemental \autoref{fig:wm_v_dd})). 
Such models have been used extensively in ecological modeling
\citep{durrett1994importance,bolker1997using,law2003population,fournier2004microscopic,champer2019suppression}
but rarely in population genetics, 
where to our knowledge implementations of continuous space models
before their availability through SLiM \citep{Haller2019}
have focused on a small number of genetic loci
\citep[e.g.,][]{slatkin1989comparison,barton2002neutral,robledoarnuncio2010isolation,rossine2014espao,jackson2014landscape},
which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. 
By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.

@@

To maintain a stable population, mortality increases with local population density.
To do this we say that individuals at distance $d$ have a competitive interaction with strength $g(d)$.
Then, the sum of all competitive interactions with individual $i$ is $n_i = \sum_j g(d_{ij})$,
where $d_{ij}$ is the distance between individuals $i$ and $j$ and the sum is over all neighbors within distance $3 \sigma$.
Since $g$ is a probability density, $n_i$ is an estimate of the number of nearby individuals per unit area.
Then, given a per-unit carrying capacity $K$, \llabel{ll:K_defn}
the probability of survival until the next time step for individual $i$ is
\begin{equation} \label{eqn:p_survival}
    p_i 
    =
    \min\left( 0.95, \frac{1}{1 + n_i / (K (1+L))} \right) .
\end{equation}
We chose this functional form so that the equilibrium population density per unit area is close to $K$, \llabel{ll:K_again}
and the mean lifetime is around $L$;
for more description see the Appendix. % \ref{apx:demographic_model}.

@@

Our demographic model includes parameters that control population density ($K$), 
mean life span ($L$), and dispersal distance ($\sigma$).
However, nonlinearity of local demographic stochasticity causes
actual realized averages of these demographic quantities to deviate from the specified values
in a way that depends on the neighborhood size.
Therefore, to properly compare to theoretical expectations,
we empirically calculated these demographic quantities in simulations. 
We recorded the census population size in all simulations,
and used mean population density ($\rho$, census size divided by total area)
to compute neighborhood size as $N_W = 4 \pi \rho \sigma^2$.  \llabel{ll:nw_defn}
To estimate generation times, we stored ages of the parents of every new individual born across 200 timesteps, 
after a 100 generation burn-in, and took the mean. 
To estimate variance in offspring number, 
we tracked the lifetime total number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period,
% subset the resulting table to include only the last timestep recorded for each individual,
and calculated the variance in number of offspring across all individuals in timesteps 50-100.
All calculations were performed with information recorded in the tree sequence,
using pyslim \citep{pyslim}.

 Note that $\sigma$ controls the dispersal of offspring away from only \emph{one} parent \llabel{ll:sigma_effective}
(e.g., the seed parent for plants),
while population genetics usually defines "dispersal distance" to be the distance to a randomly chosen one of the two parents
(thus taking into account the distance from pollen parent as well).
This second component -- the distance between mates --
has in our simulations a distribution that is of order $\sigma$ but that depends on the population's patchiness.
If both between-mate distance and dispersal distance has variance $\sigma^2$ along each axis,
then the mean squared distance to a randomly chosen parent along that axis would be
$(\sigma^2 + 2 \sigma^2)/2 = 3 \sigma^2 /2$.
To match theory, neighborhood size should be defined in terms of \emph{effective}
dispersal distance, i.e., the mean squared displacement along an axis
between parent-child pairs found moving back along a lineage \citep{barton2002neutral}, 
and \emph{effective} population density \citep{Rousset1997}.
However, we use $\sigma$ and $\rho$ as defined here to compute $N_W$
because these quantities are more easily observable in practice than their ``effective'' versions.

@@

\begin{figure}[htbp]
\centering
\includegraphics{figures/sampling_maps.pdf}
\caption{Example sampling maps for 60 individuals on a 50$\times$50 landscape for midpoint, point, and random sampling strategies, respectively.}
\label{fig:samplemap}
\end{figure}

%%%%%%%%%%%%%
\subsection{Summary Statistics}
We calculated the site frequency spectrum and a set of 18 summary statistics (Table \ref{table:sumstats}) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}. 
Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{IS}$), and Tajima's $D$, 
as well as (motivated by \citet{Rousset1997}'s results)
the correlation coefficient between the logarithm of the spatial distance and the proportion of identical base pairs across pairs of individuals. 
% as an isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distance analogous to \citet{Rousset1997}'s approach, which we summarized as

Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep[e.g.,][]{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among sampled chromosomes,
defined to be the set of distances between adjacent sites that differ between the two chromosomes. 
The full distribution of lengths of IBS tracts for each pair of chromosomes was first calculated with a custom python function. 
%\url{https://github.com/cjbattey/spaceness/blob/master/scripts/slimtools.py}. 
We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $10^6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons. 
We then calculated correlation coefficients between spatial distance and each moment of the pairwise IBS tract distribution. Because more closely related individuals on average share longer haplotype blocks we expect that spatial distance will be negatively correlated with mean haplotype block length, and that this correlation will be strongest (i.e., most negative) when dispersal is low. The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks, and so should reflect recent demographic events \citep{chapman2002effect}. For a subset of simulations, we also calculated cumulative distributions for IBS tract lengths across pairs of distant (more than 40 map units) and nearby (less than 10 map units) individuals. Last, we examined the relationship between allele frequency and the spatial dispersion of an allele by calculating the average distance between individuals carrying each derived allele.

@@

To evaluate the performance of these methods we binned simulations by neighborhood size, took a rolling median of inferred $N_{e}$ trajectories across all model fits in a bin for each method and sampling strategy.  We also examined how varying levels of isolation by distance impacted the variance of $N_{e}$ estimates by calculating the standard deviation of $N_{e}$ from each best-fit model. % and plotting these against neighborhood size.

%In preliminary runs we found that inferred population histories were highly variable even when simulating under a coalescent model, suggesting that some of the differences in demographic estimates for spatial models are caused by the behavior of the optimization algorithm rather than bias in the SFS caused by spatial mate choice and dispersal. To separate these effects we ran 100 coalescent simulations with constant population size $6.1\times 10^{-3}$ (the mean $N_{e}$ of random-mating SLiM models estimated from $\Theta_{\pi}$) and fit stairwayplot models using the same script as for our spatial models. All coalescent simulations were performed using msprime \citep{Kelleher2016}. We then calculated the standard deviation of inferred $N_{e}$ in each stairwayplot model to summarize the degree of fluctuation around the simulated population size, and asked if standard deviations were higher in spatial relative to coalescent models with a one-tailed t-test.

@@

Adjusting the spatial dispersal and interaction distance, $\sigma$,
has a surprisingly large effect on demographic quantities that are usually fixed in Wright-Fisher models --
the generation time, census population size, and variance in offspring number, shown in Figure~\ref{fig:genparams}. Because our simulation is paramaterized on an individual level, these population parameters emerge as a property of the interactions among individuals rather than being directly set. \llabel{emergentparams}
Variation across runs occurs because,
even though the parameters $K$ and $L$ that control population density and mean lifetime respectively \llabel{ll:K_more} 
were the same in all simulations,
the strength of stochastic effects depends strongly on the spatial interaction distance $\sigma$.
For instance, 
the population density near to individual $i$ (denoted $n_i$ above)
is computed by averaging over roughly $N_W = 4 \pi K \sigma^2$ individuals,
and so has standard deviation proportional to $1/\sqrt{N_W}$ --
it is more variable at lower densities.
(Recall that $N_W$ is Wright's neighborhood size.)
Since the probability of survival is a nonlinear function of $n_i$,
actual equilibrium densities and lifetimes differ from $K$ and $L$.
This is the reason that we included \emph{random mating} simulations --
where mate choice and offspring dispersal are both nonspatial --
since this should preserve the random fluctuations in local population density
while destroying any spatial genetic structure.
We verified that random mating models retained no geographic signal
by showing that summary statistics did not differ significantly between sampling regimes
(Table~\ref{table:sampling}), unlike in spatial models (discussed below).

There are a few additional things to note about Figure~\ref{fig:genparams}. 
First, all three quantities are non-monotone with neighborhood size. 
Census size largely declines as neighborhood size increases for both the spatial and random mating models. 
However, for spatial models this decline only begins for neighborhood size $\geq 10$. 
Spatial and random mating models are indistinguishable from one another
for neighborhood sizes larger than 100.
Census sizes range from around 14,000 at low $\sigma$ in the random mating model to 10,000 for both models when neighborhood sizes approach 1,000.
The scaling of census sizes in both random-mating and spatial models appears to be related to two consequences of the spatial competition function: the decline of fitness at range edges, which effectively reduces the habitable area by one $\sigma$ around the edge of the map and so results in a smaller habitable area at high $\sigma$ values; and variation in the equilibrium population density given varying competition radii.
Furthermore, census size increases in spatial models as neighborhood size increases from 2 to 10.
This may reflect an Allee effect \citep{Allee1949} in which some individuals are unable to find mates when the mate selection radius is very small.  \llabel{censusscaling}

\begin{figure}[htbp]
\centering
\includegraphics{figures/pop_params.pdf}
\caption{Genealogical parameters from spatial and random mating SLiM simulations, by neighborhood size.}
\label{fig:genparams}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{figures/runtimes.pdf}
\caption{Run times of continuous space simulations with landscape width 50 and expected density 5 under varying neighborhood size. Times are shown for simulations run with mutations applied directly in SLiM (dashed lines) or later applied to tree sequences with msprime (solid lines). Times for simulations run with tree sequence recording disabled are shown in grey.}
\label{fig:runtimes}
\end{figure}

@@

Run times for our model scale approximately linearly with neighborhood size (Figure \ref{fig:runtimes}), with the lowest neighborhood sizes reaching 30N generations in around a day and those with neighborhood size approaching 1,000 requiring up to three weeks of computation. As currently implemented, running simulations at neighborhood sizes more than 1,000 to coalescence is likely impractical, though running these models for more limited timescales and then ``recapitating'' the simulation using reverse-time simulation from the resulting tree sequence in msprime is possible \citep{haller2019treesequence}.  \llabel{runtimesxn}

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/sfs_w_sumstats_loglog.pdf}
\caption{Site frequency spectrum (A; note axes are log-scaled) and summary statistic distributions (B) by sampling strategy and neighborhood size.}
\label{fig:sumstats}
\end{figure*}

@@

\begin{figure*}[p]
\centering
\includegraphics[width=.5\textwidth]{figures/ibs_cumdists_w_rm.pdf}
\caption{
    Cumulative distributions for IBS tract lengths per pair of individuals at different geographic distances, across three neighborhood sizes (NS).
    Nearby pairs (red curves) share many more long IBS tracts than do distant pairs (blue curves),
    except in the random mating model.
    The distribution of long IBS tracts between nearby individuals are very similar across neighborhood sizes,
    but distant individuals are much more likely to share long IBS tracts at high neighborhood size
    than at low neighborhood size.
}
\label{fig:ibs}
\end{figure*}

@@

%%%%%%%%%%%
\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/distance_by_DAF.pdf}
\caption{
    Spatial spread of rare alleles by neighborhood size (NS):
    Each plot shows the distribution (across derived alleles and simulations)
    of average pairwise distance between individuals carrying a focal derived allele
    and derived allele frequency.
}
\label{fig:dafdists}
\end{figure*}

@@

\begin{figure}[p]
\centering
\includegraphics[width=\textwidth]{figures/demography_rollmedian_w_variance.pdf}
\caption{A: Rolling median inferred $N_{e}$ trajectories for stairwayplot and smc++ across sampling strategies and neighborhood size bins. The dotted line shows the mean $N_{e}$ of random-mating simulations. B: Standard deviation of individual inferred $N_{e}$ trajectories, by neighborhood size and sampling strategy. Black lines are loess curves. Plots including individual model fits are shown in Figure \ref{fig:demography_supplement}.}
\label{fig:demography}
\end{figure}

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/gwas_summary_loglog_qqlines.pdf}
    \caption{Impacts of spatially varying environments and isolation by distance on linear regression GWAS. Simulated quantitative phenotypes are determined only by an individual's location and the spatial distribution of environmental factors. In \textbf{A} we show the phenotypes and locations of sampled individuals under four environmental distributions, with transparency scaled to phenotype. As neighborhood size increases a PCA explains less of the total variation in the data (\textbf{B}).
    Spatially correlated environmental factors cause false positives at a large proportion of SNPs, which is partially but not entirely corrected by adding the first 10 PC coordinates as covariates (\textbf{C}).
    Quantile-quantile plots in (\textbf{D}) show inflation of $-\log_{10}(p)$ after PC correction for simulations with spatially structured environments, with line colors showing the neighborhood size of each simulation.
}
\label{fig:gwas}
\end{figure*}

@@

Spatial structure is particularly challenging for genome-wide association studies, because the effects of dispersal on genetic variation are compounded by spatial variation in the environment \citep{Mathieson2012}. Spatially restricted mate choice and dispersal causes variation in allele frequencies across the range of a species. If environmental factors affecting the phenotype of interest also vary over space, then allele frequencies and environmental exposures will covary over space. \llabel{ll:1:11:2}
In this scenario an uncorrected GWAS will infer genetic associations with a purely environmental phenotype at any site in the genome that is differentiated over space, and the relative degree of bias will be a function of the degree of covariation in allele frequencies and the environment (i.e., Figure \ref{fig:gwas}C, bottom panel). This pattern has been demonstrated in a variety of simulation and empirical contexts \citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015,Berg2018,Sohail2018}.

@@

\bibliography{spaceness,more_refs}

@@

%%%%%%%%%%%%%%%%%%%%%%%
\section{Comparisons with Stepping-Stone Models}
\llabel{apx:stepping_stone}

@@

\includereviews

>>>>

%%%%%%%%%%%
% for responses to reviewers
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\input{review-response-commands}
\newcommand{\responsefile}{second-review-responses.tex}  % name of the review reponses file

%Space is the Place: How Dispersal and Competition Shape Genetic Variation in Continuous Space
%Impacts of Continuous Spatial Structure on Analyses of Population Genetic Data
%Impacts of Isolation by Distance on Demographic Modeling and Association Studies
%Space is the Place: How Dispersal and Sampling Shape Inference From Genetic Variation in Continuous Space
%Evolution on the Final Frontier: How Dispersal and Sampling Shape Inference From Genetic Data in Continuous Space
\title{Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data}

@@

%% For the footnote.
%% Give the last name of the first author if only one author;
% \runningauthor{FirstAuthorLastname}
%% last names of both authors if there are two authors;
% \runningauthor{FirstAuthorLastname and SecondAuthorLastname}
%% last name of the first author followed by et al, if more than two authors.
\runningauthor{Battey \textit{et al.}}

@@

The assumption that populations are ``well-mixed'' has important implications for downstream inference of selection and demography. Methods based on the coalescent \citep{Kingman1982,wakeley2005coalescent} assume that the sampled individuals are a random draw from a well-mixed population
that is much larger than the sample \citep{wakeley2003gene}. The key assumption is that the individuals of each generation are \emph{exchangeable}, so that there is no correlation between the fate or fecundity of a parent and that of their offspring \citep{Huillet2011}. If dispersal or mate selection is limited by geographic proximity, this assumption can be violated in many ways. For instance, if mean viability or fecundity is spatially autocorrelated, then limited geographic dispersal will lead to parent--offspring correlations. Furthermore, nearby individuals will be more closely related than an average random pair, so drawing multiple samples from the same area of the landscape will represent a biased sample of the genetic variation present in the whole population \citep{Stadler2009}.

@@

The degree to which genetic relationships are geographically correlated
depends on the chance that two geographically nearby individuals are close relatives
-- in modern terms, by the tension between
migration (the chance that one is descended from a distant location)
and coalescence (the chance that they share a parent).
A key early observation by \citet{Wright1946} is that this balance is often nicely summarized by the ``neighborhood size'',
defined in two dimensions to be $N_W = 4 \pi \rho \sigma^2$,
where $\sigma^2$ is one half of the mean squared parent--offspring distance
and $\rho$ is population density
(see \citet{Rousset1997} for further discussion of parameter definitions in one- and two-dimensional habitats). 
This can be thought of as proportional to the average number of potential mates for an individual (those within distance $2 \sigma)$,
or the number of potential parents of a randomly chosen individual. Empirical estimates of neighborhood size vary hugely across species
-- even in human populations,
estimates range from 40 to over 5,000 depending on the population and method of estimation (Table~\ref{table:NStable}).

@@

To avoid hard-to-evaluate approximations, we here used forward-time, individual-based simulations
across continuous geographical space.
The question of what regulates real populations has a long history and many answers
\citep[e.g.,][]{lloyd1967crowding,antonovics1980ecological,crawley1990population}, 
but it is clear that populations must at some point have density-dependent feedback on population size,
or else they would face eventual extinction or explosion.
In the absence of unrealistic global population regulation,
this regulation must be local,
and there are many ways to achieve this \citep{bolker2003spatial}.
In our simulations,
each individual's probability of survival is a decreasing function of local population density,
which shifts reproductive output towards low-density regions,
and produces total census sizes that fluctuate around an equilibrium. 
This also prevents the population clumping seen by \citet{Felsenstein1975}
(Supplemental \autoref{fig:wm_v_dd})). 
Such models have been used extensively in ecological modeling
\citep{durrett1994importance,bolker1997using,law2003population,fournier2004microscopic,champer2019suppression}
but rarely in population genetics, 
where to our knowledge implementations of continuous space models
before their availability through SLiM \citep{Haller2019}
have focused on a small number of genetic loci
\citep[e.g.,][]{slatkin1989comparison,barton2002neutral,robledoarnuncio2010isolation,rossine2014espao,jackson2014landscape},
which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. 
By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.

@@

To maintain a stable population, mortality increases with local population density.
To do this we say that individuals at distance $d$ have a competitive interaction with strength $g(d)$.
Then, the sum of all competitive interactions with individual $i$ is $n_i = \sum_j g(d_{ij})$,
where $d_{ij}$ is the distance between individuals $i$ and $j$ and the sum is over all neighbors within distance $3 \sigma$.
Since $g$ is a probability density, $n_i$ is an estimate of the number of nearby individuals per unit area.
Then, given a per-unit carrying capacity $K$, 
the probability of survival until the next time step for individual $i$ is
\begin{equation} \label{eqn:p_survival}
    p_i 
    =
    \min\left( 0.95, \frac{1}{1 + n_i / (K (1+L))} \right) .
\end{equation}
We chose this functional form so that the equilibrium population density per unit area is close to $K$, 
and the mean lifetime is around $L$;
for more description see the Appendix.

@@

Our demographic model includes parameters that control population density ($K$), 
mean life span ($L$), and dispersal distance ($\sigma$).
However, nonlinearity of local demographic stochasticity causes
actual realized averages of these demographic quantities to deviate from the specified values
in a way that depends on the neighborhood size.
Therefore, to properly compare to theoretical expectations,
we empirically calculated these demographic quantities in simulations. 
We recorded the census population size in all simulations,
and used mean population density ($\rho$, census size divided by total area)
to compute neighborhood size as $N_W = 4 \pi \rho \sigma^2$.  
To estimate generation times, we stored ages of the parents of every new individual born across 200 timesteps, 
after a 100 generation burn-in, and took the mean. 
To estimate variance in offspring number, 
we tracked the lifetime total number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period,
and calculated the variance in number of offspring across all individuals in timesteps 50-100.
All calculations were performed with information recorded in the tree sequence,
using pyslim \citep{pyslim}.

 Note that $\sigma$ controls the dispersal of offspring away from only \emph{one} parent (e.g., the seed parent for plants),
while population genetics usually defines "dispersal distance" to be the distance to a randomly chosen one of the two parents
(thus taking into account the distance from pollen parent as well).
This second component -- the distance between mates --
has in our simulations a distribution that is of order $\sigma$ but that depends on the population's patchiness.
If both between-mate distance and dispersal distance has variance $\sigma^2$ along each axis,
then the mean squared distance to a randomly chosen parent along that axis would be
$(\sigma^2 + 2 \sigma^2)/2 = 3 \sigma^2 /2$.
To match theory, neighborhood size should be defined in terms of \emph{effective}
dispersal distance, i.e., the mean squared displacement along an axis
between parent-child pairs found moving back along a lineage \citep{barton2002neutral}, 
and \emph{effective} population density \citep{Rousset1997}.
However, we use $\sigma$ and $\rho$ as defined here to compute $N_W$
because these quantities are more easily observable in practice than their ``effective'' versions.

@@

\begin{figure}[htbp]
\centering
\includegraphics{final_figures/Fig1_sampling_maps.pdf}
\caption{Example sampling maps for 60 individuals on a 50$\times$50 landscape for midpoint, point, and random sampling strategies, respectively.}
\label{fig:samplemap}
\end{figure}

%%%%%%%%%%%%%
\subsection{Summary Statistics}
We calculated the site frequency spectrum and a set of 18 summary statistics (Table \ref{table:sumstats}) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}. 
Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{IS}$), and Tajima's $D$, 
as well as (motivated by \citet{Rousset1997}'s results)
the correlation coefficient between the logarithm of the spatial distance and the proportion of identical base pairs across pairs of individuals.

Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep[e.g.,][]{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among sampled chromosomes,
defined to be the set of distances between adjacent sites that differ between the two chromosomes. 
The full distribution of lengths of IBS tracts for each pair of chromosomes was first calculated with a custom python function. 
We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $10^6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons. 
We then calculated correlation coefficients between spatial distance and each moment of the pairwise IBS tract distribution. Because more closely related individuals on average share longer haplotype blocks we expect that spatial distance will be negatively correlated with mean haplotype block length, and that this correlation will be strongest (i.e., most negative) when dispersal is low. The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks, and so should reflect recent demographic events \citep{chapman2002effect}. For a subset of simulations, we also calculated cumulative distributions for IBS tract lengths across pairs of distant (more than 40 map units) and nearby (less than 10 map units) individuals. Last, we examined the relationship between allele frequency and the spatial dispersion of an allele by calculating the average distance between individuals carrying each derived allele.

@@

To evaluate the performance of these methods we binned simulations by neighborhood size, took a rolling median of inferred $N_{e}$ trajectories across all model fits in a bin for each method and sampling strategy.  We also examined how varying levels of isolation by distance impacted the variance of $N_{e}$ estimates by calculating the standard deviation of $N_{e}$ from each best-fit model.

@@

Adjusting the spatial dispersal and interaction distance, $\sigma$,
has a surprisingly large effect on demographic quantities that are usually fixed in Wright-Fisher models --
the generation time, census population size, and variance in offspring number, shown in Figure~\ref{fig:genparams}. Because our simulation is paramaterized on an individual level, these population parameters emerge as a property of the interactions among individuals rather than being directly set. 
Variation across runs occurs because,
even though the parameters $K$ and $L$ that control population density and mean lifetime respectively  
were the same in all simulations,
the strength of stochastic effects depends strongly on the spatial interaction distance $\sigma$.
For instance, 
the population density near to individual $i$ (denoted $n_i$ above)
is computed by averaging over roughly $N_W = 4 \pi K \sigma^2$ individuals,
and so has standard deviation proportional to $1/\sqrt{N_W}$ --
it is more variable at lower densities.
(Recall that $N_W$ is Wright's neighborhood size.)
Since the probability of survival is a nonlinear function of $n_i$,
actual equilibrium densities and lifetimes differ from $K$ and $L$.
This is the reason that we included \emph{random mating} simulations --
where mate choice and offspring dispersal are both nonspatial --
since this should preserve the random fluctuations in local population density
while destroying any spatial genetic structure.
We verified that random mating models retained no geographic signal
by showing that summary statistics did not differ significantly between sampling regimes
(Table~\ref{table:sampling}), unlike in spatial models (discussed below).

There are a few additional things to note about Figure~\ref{fig:genparams}. 
First, all three quantities are non-monotone with neighborhood size. 
Census size largely declines as neighborhood size increases for both the spatial and random mating models. 
However, for spatial models this decline only begins for neighborhood size $\geq 10$. 
Spatial and random mating models are indistinguishable from one another
for neighborhood sizes larger than 100.
Census sizes range from around 14,000 at low $\sigma$ in the random mating model to 10,000 for both models when neighborhood sizes approach 1,000.
The scaling of census sizes in both random-mating and spatial models appears to be related to two consequences of the spatial competition function: the decline of fitness at range edges, which effectively reduces the habitable area by one $\sigma$ around the edge of the map and so results in a smaller habitable area at high $\sigma$ values; and variation in the equilibrium population density given varying competition radii.
Furthermore, census size increases in spatial models as neighborhood size increases from 2 to 10.
This may reflect an Allee effect \citep{Allee1949} in which some individuals are unable to find mates when the mate selection radius is very small.

\begin{figure}[htbp]
\centering
\includegraphics{final_figures/Fig2_pop_params.pdf}
\caption{Genealogical parameters from spatial and random mating SLiM simulations, by neighborhood size.}
\label{fig:genparams}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{final_figures/Fig3_runtimes.pdf}
\caption{Run times of continuous space simulations with landscape width 50 and expected density 5 under varying neighborhood size. Times are shown for simulations run with mutations applied directly in SLiM (dashed lines) or later applied to tree sequences with msprime (solid lines). Times for simulations run with tree sequence recording disabled are shown in grey.}
\label{fig:runtimes}
\end{figure}

@@

Run times for our model scale approximately linearly with neighborhood size (Figure \ref{fig:runtimes}), with the lowest neighborhood sizes reaching 30N generations in around a day and those with neighborhood size approaching 1,000 requiring up to three weeks of computation. As currently implemented, running simulations at neighborhood sizes more than 1,000 to coalescence is likely impractical, though running these models for more limited timescales and then ``recapitating'' the simulation using reverse-time simulation from the resulting tree sequence in msprime is possible \citep{haller2019treesequence}.

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{final_figures/Fig4_sfs_w_sumstats_loglog.pdf}
\caption{Site frequency spectrum (A; note axes are log-scaled) and summary statistic distributions (B) by sampling strategy and neighborhood size.}
\label{fig:sumstats}
\end{figure*}

@@

\begin{figure*}[p]
\centering
\includegraphics[width=.5\textwidth]{final_figures/Fig5_ibs_cumdists_w_rm.pdf}
\caption{
    Cumulative distributions for IBS tract lengths per pair of individuals at different geographic distances, across three neighborhood sizes (NS).
    Nearby pairs (red curves) share many more long IBS tracts than do distant pairs (blue curves),
    except in the random mating model.
    The distribution of long IBS tracts between nearby individuals are very similar across neighborhood sizes,
    but distant individuals are much more likely to share long IBS tracts at high neighborhood size
    than at low neighborhood size.
}
\label{fig:ibs}
\end{figure*}

@@

%%%%%%%%%%%
\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{final_figures/Fig6_distance_by_DAF.pdf}
\caption{
    Spatial spread of rare alleles by neighborhood size (NS):
    Each plot shows the distribution (across derived alleles and simulations)
    of average pairwise distance between individuals carrying a focal derived allele
    and derived allele frequency.
}
\label{fig:dafdists}
\end{figure*}

@@

\begin{figure}[p]
\centering
\includegraphics[width=\textwidth]{final_figures/Fig7_demography_rollmedian_w_variance.pdf}
\caption{A: Rolling median inferred $N_{e}$ trajectories for stairwayplot and smc++ across sampling strategies and neighborhood size bins. The dotted line shows the mean $N_{e}$ of random-mating simulations. B: Standard deviation of individual inferred $N_{e}$ trajectories, by neighborhood size and sampling strategy. Black lines are loess curves. Plots including individual model fits are shown in Figure \ref{fig:demography_supplement}.}
\label{fig:demography}
\end{figure}

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{final_figures/Fig8_gwas_summary_loglog_qqlines.pdf}
    \caption{Impacts of spatially varying environments and isolation by distance on linear regression GWAS. Simulated quantitative phenotypes are determined only by an individual's location and the spatial distribution of environmental factors. In \textbf{A} we show the phenotypes and locations of sampled individuals under four environmental distributions, with transparency scaled to phenotype. As neighborhood size increases a PCA explains less of the total variation in the data (\textbf{B}).
    Spatially correlated environmental factors cause false positives at a large proportion of SNPs, which is partially but not entirely corrected by adding the first 10 PC coordinates as covariates (\textbf{C}).
    Quantile-quantile plots in (\textbf{D}) show inflation of $-\log_{10}(p)$ after PC correction for simulations with spatially structured environments, with line colors showing the neighborhood size of each simulation.
}
\label{fig:gwas}
\end{figure*}

@@

Spatial structure is particularly challenging for genome-wide association studies, because the effects of dispersal on genetic variation are compounded by spatial variation in the environment \citep{Mathieson2012}. Spatially restricted mate choice and dispersal causes variation in allele frequencies across the range of a species. If environmental factors affecting the phenotype of interest also vary over space, then allele frequencies and environmental exposures will covary over space. 
In this scenario an uncorrected GWAS will infer genetic associations with a purely environmental phenotype at any site in the genome that is differentiated over space, and the relative degree of bias will be a function of the degree of covariation in allele frequencies and the environment (i.e., Figure \ref{fig:gwas}C, bottom panel). This pattern has been demonstrated in a variety of simulation and empirical contexts \citep{Price2006,Yu2006,Young2018,Mathieson2012,Kang2008,Kang2010,Bulik-Sullivan2015,Berg2018,Sohail2018}.

@@

\bibliography{spaceness_final}

@@

%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\section{Appendix 1}

\subsection{Comparisons with Stepping-Stone Models}

@@
